Different forms of uncertainty differently affect the evolution of social learning

### 2022-07-16 ms.tex
revised Model section intro paragraphs; next: revise details of each model component

Despite the centrality of temporal fluctuations, even this construct is implemented in several ways \cite{aoki2014evolution}. Most commonly, it is modeled as an independent probability of the environment changing its state each generation ~\cite{BoydRicherson1985,Rogers1988,Feldman1996,McElreath2005,Enquist2007,perreault2012bayesian,aoki2014evolution}. The probabilities of environmental change are usually fixed per generation, %~\cite{BoydRicherson1985,Rogers1988,kendal2009evolution}, 
but they have also been modeled as deterministic cycles ~\cite{Feldman1996, aoki2014evolution}. Furthermore, the consequences of environmental change can range from mild to  catastrophic. In the latter case, a change of environment results in the total elimination of any adaptive behavior, which must be learned \emph{de novo} for the new environment ~\cite{Rogers1988}. Such catastrophic environmental changes present a large adaptive challenge as individuals cannot bank on accumulated information from previous generations. Other models of environmental change introduce less uncertainty as they fluctuate between two environmental states with corresponding adaptive behaviors. This means that all individuals with previously maladaptive behaviors then have adaptive behaviors upon the environment changing ~\cite{perreault2012bayesian}. Which strategy one chooses for modelling temporal change can have consequences for at least some of our inferences  ~\cite{aoki2014evolution}.

@@

\textbf{Payoff ambiguity.}
Many models of social learning differentiate between the payoffs for adopting optimal vs. non-optimal behaviors \cite{Rogers1988,Enquist2007,Rendell2010}. The size of the difference between these payoffs is usually taken to indicate the strength of selection of learning, which makes sense. However, in reality payoffs for particular behaviors are not always consistent. A behavior may yield a large payoff sometimes and a small payoff other times \cite{McElreath2005}. This means that signals about the relationship between behavior and payoff are often noisy, and differentiating between behavioral options is in part a problem of signal detection.
When the difference in expected payoffs between optimal and non-optimal behaviors is very large, this noise matters little, as the signal is still very clear. However, when the expected payoffs of different behaviors are similar relative to the size of their variances, ambiguity arises about which behaviors really are superior. Smaller differences between payoffs corresponds to larger ambiguity. Payoff ambiguity has been manipulated in both theoretical ~\cite{perreault2012bayesian} and empirical ~\cite{McElreath2005l, Morgan2012} studies, both of which support the claim that payoff ambiguity increases the reliance on social information. Importantly, payoff ambiguity %differs from selection set size in that it not only affects uncertainty, it also directly
influences both uncertainty \emph{and} the strength of selection. %This parameter sometimes takes the form of differences in expected payoffs arising from different behaviors or strategies~\cite{Enquist2007,Rendell2010} or by varying the standard deviation of payoffs~\cite{McElreath2005}.

@@

\cm{Other choices we made that might need to be justified: 1) softmax: existing models underestimate the power of learning from experience by not having a rich cognition. This rich cognition can improve both individual learning and help social learners recover from outdated / misleading social info. 2) no conformism: but that's ok because we're giving social learning the best chance of success. 3) no horizontal transmission, no spatial variation}

\begin{table}[h]
    \caption{Dynamic agent-level variables. Each has an implicit time dependence.}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{4.0in}p{1.25in}} \toprule

@@

        $\pi_i$ & Net payoffs accumulated by $i$ within generation
                                & 0.0 \\
        \bottomrule
    \end{tabular}
\end{table}

To answer our research questions we developed an agent-based model of social
learning evolution in which a society of $N$ individuals \cm{do we ever define N used?} each must decide which of
$B$ behaviors to perform at each time step within a generation of $L$ time steps. At
the end of each generation, agents are selected with replacement from the population
to reproduce another $N$ agents. Agents who inherit the social learning trait then
learn about behavioral payoffs from the previous generation, while asocial learners
begin life with no \emph{a priori} knowledge of the world.  Each behavior is a
``bandit'', a common modeling and experimental approach for representing behaviors
with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Steyvers2009,Rendell2010,Schulz2019}.

We identified four ``principal components'' of uncertainty that affect the
ability of agents to find and choose the optimal behavior: 
\emph{environmental variability}; \emph{payoff ambiguity}; \emph{selection set size}
(number of behaviors environment affords); and the \emph{effective lifespan}.
Environmental variability, denoted $u$, is the probability that the behavior
yielding the optimal payoff, $\pihigh$, changes from one generation to the next.
We operationalized payoff ambiguity by setting $\pihigh = 0.9$ for all analyses
and varying $\pilow \in \{0.1, 0.45, 0.8\}$. The selection set size is the
total number of possible behaviors agents choose from, denoted $B$. The 
effective life span is the number of time steps agents live for, which is
also the number of behaviors each agent performs each generation.

\vspace{2em}
\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.
      Bold indicates \emph{default value tested}.
    \mt{TODO: Test more $N_\text{T}$ for sensitivity analysis}}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{4.0in}p{1.25in}} \toprule

        $B$       & Number of possible behaviors (represented by ``bandits'') 
                  & 2, 4, 10 \\

@@

        $L$    & Number of time steps per generation & 1, $B/2$, $B$, $2B$, $4B$ \\

        $N_T$    & Number of teachers to pool, from which best selected 
                 & \textbf{5}  \\

@@

All agents perform individual, trial-and-error learning over their lifespan.
This learning is guided based on undirected, softmax search which drives
agents to explore all behavioral options in the selection set when 
the agent is more uncertain which is optimal, and to exploit the most profitable
behaviors when the agent is more certain which are most beneficial.

@@

\subsubsection{Initialization}

@@

@@

By identifying four common uncertainty dimensions we developed a model that predicts
how different forms of uncertainty might interact to foster or suppress social
learning.  The model reproduced classic predictions that social learning can yield
higher payoffs than individual learning when uncertainty is sufficiently limited.
By disambiguating and organizing various operationalizations of uncertainty, 
we developed a more nuanced theoretical understanding of which forms of
uncertainty interact in which ways to make social learning more or less beneficial
compared to asocial learning.  Note that in many cases we made modeling choices that
excluded some important mechanisms of social learning; for example we assumed
success-biased learning, although conformity is theoretically important as
well~\cite{Muthukrishna2016a,Smaldino2018b}. Still, we believe our model
advances theoretical parsimony; and the software implementation is designed to
facilitate testing alternate modeling choices.

% \paragraph{Disc. 2: Foundation for studying social learning evolution with group
% structure or other topics(PAUL, JAMIE, CRISTINA)}~\cite{Katsnelson2014}.

Our modeling approach may have applications to developing social artificial
intelligence. Much effort has gone in to developing the cognitive machinery to
realize multi-agent reinforcement learning
systems~\cite{Sandholm1996,Ndousse2021,Gronauer2022}, with some work explicitly
incorporating cultural evolutionary approaches to social learning~\cite{Jaques2019}.
However, it seems there are a lack of methods for adapting to environmental
uncertainty, or determining when social learning is actually advantageous under
different uncertainty conditions. Our model dynamics could be used as an
evolutionary algorithm to update AI agents' social learning strategy without
explicit knowledge of, or need to approximate, the underlying uncertainty structure.
Our model could also predict the expected benefit of social learning, which may be
inherently costly in artificial multi-agent learning systems, and how long the
translated evolutionary algorithm may take to converge.

Taking the functional form of uncertainty seriously is critical for understanding how humans adapt to an uncertain and rapidly changing world, both to mitigate existential threats~\cite{Moya2020,Jones2021}, and to capitalize on new opportunities such as transitioning to clean energy use and production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

\bibliographystyle{apacite}
% \bibliography{/Users/mt/workspace/Writing/library.bib}
\bibliography{this.bib}

>>>>

Despite the centrality of temporal fluctuations, even this construct is implemented in several ways \cite{aoki2014evolution}. Most commonly, it is modeled as an independent probability of the environment changing its state each generation ~\cite{BoydRicherson1985,Rogers1988,Feldman1996,McElreath2005,Enquist2007,perreault2012bayesian,aoki2014evolution}. The probabilities of environmental change are usually fixed per generation, %~\cite{BoydRicherson1985,Rogers1988,kendal2009evolution}, 
but they have also been modeled as deterministic cycles ~\cite{Feldman1996, aoki2014evolution}. Furthermore, the consequences of environmental change can range from mild to  catastrophic. In the latter case, a change of environment results in the total elimination of any adaptive behavior, which must be learned \emph{de novo} for the new environment~\cite{Rogers1988}. Such catastrophic environmental changes present a large adaptive challenge as individuals cannot bank on accumulated information from previous generations. Other models of environmental change introduce less uncertainty as they fluctuate between two environmental states with corresponding adaptive behaviors. This means that all individuals with previously maladaptive behaviors then have adaptive behaviors upon the environment changing ~\cite{perreault2012bayesian}. Which strategy one chooses for modelling temporal change can have consequences for at least some of our inferences  ~\cite{aoki2014evolution}.

@@

\textbf{Payoff ambiguity.}
Many models of social learning differentiate between the payoffs for adopting optimal vs. non-optimal behaviors \cite{Rogers1988,Enquist2007,Rendell2010}. The size of the difference between these payoffs is usually taken to indicate the strength of selection of learning, which makes sense. However, in reality payoffs for particular behaviors are not always consistent. A behavior may yield a large payoff sometimes and a small payoff other times \cite{McElreath2005}. This means that signals about the relationship between behavior and payoff are often noisy, and differentiating between behavioral options is in part a problem of signal detection.
When the difference in expected payoffs between optimal and non-optimal behaviors is very large, this noise matters little, as the signal is still very clear. However, when the expected payoffs of different behaviors are similar relative to the size of their variances, ambiguity arises about which behaviors really are superior. Smaller differences between payoffs corresponds to larger ambiguity. Payoff ambiguity has been manipulated in both theoretical ~\cite{perreault2012bayesian} and empirical ~\cite{McElreath2005, Morgan2012} studies, both of which support the claim that payoff ambiguity increases the reliance on social information. Importantly, payoff ambiguity %differs from selection set size in that it not only affects uncertainty, it also directly
influences both uncertainty \emph{and} the strength of selection. %This parameter sometimes takes the form of differences in expected payoffs arising from different behaviors or strategies~\cite{Enquist2007,Rendell2010} or by varying the standard deviation of payoffs~\cite{McElreath2005}.

@@

% \cm{Other choices we made that might need to be justified: 1) softmax: existing models underestimate the power of learning from experience by not having a rich cognition. This rich cognition can improve both individual learning and help social learners recover from outdated / misleading social info. 2) no conformism: but that's ok because we're giving social learning the best chance of success. 3) no horizontal transmission, no spatial variation}

To address our research questions we developed an evolutionary agent-based model of
social learning evolution.  Evolutionary agent-based models simulate changes in
prevalence of a phenotype over several generations---in this model the phenotype, or
``trait'', is whether or not an agent engages in social learning.  In the model, a
population of $N$ individuals each must decide which of $B$ behaviors to perform at
each time step within a generation consisting of $L$ time steps. At the end of each
generation, agents are selected with replacement, biased by accumulated payoffs
within the generation, from the population to reproduce another $N$ agents. Agents
who inherit the social learning trait then learn about behavioral payoffs from the
previous generation, while asocial learners begin life with no \emph{a priori}
knowledge of the world.  Each behavior is a ``bandit'', a common modeling and
experimental approach for representing behaviors with Bernoulli-distributed
payoffs~\cite{SuttonBartoBook,McElreath2005,Steyvers2009,Rendell2010,Schulz2019}.
All bandits thus yield a payoff of 1 or 0 with a certain probability.  One of the
bandits/behaviors in the model is more likely to pay off than all the rest.  Agents
optimize their net payoffs over their lifespan by quickly figuring out which
behavior is optimal and performing it as often as possible within their lifespan.

We operationalized the four ``principal components'' of uncertainty that affect the
ability of agents to find and choose the optimal behavior. Recall these varieties of
uncertainty are \emph{environmental variability}; \emph{payoff ambiguity};
\emph{selection set size}; and the \emph{effective lifespan}. Environmental
variability is the probability that the behavior yielding the optimal payoff changes
from one generation to the next. We do not consider cases where the optimal behavior
varies within a generation. This is empirically justified since cultural evolution
occurs when meta-cognitive strategies and information are transferred or not via
learning and social influence, instead of truly by physical reproduction and death
as in genetic evolution.  The selection set size is the total number of possible
behaviors agents choose from, assumed to be constant within a simulation.  We
operationalized payoff ambiguity by manipulating the payoff difference between optimal
and non-optimal behaviors. The effective life span is the
number of time steps in which agents perform behaviors within a generation. 
Our model is not spatial, so agents have free, instant
access to all behaviors, and there is no spatial uncertainty, 
as might be the case for humans in a computer-based lab
experiment~\cite{McElreath2005,Morgan2012}.  Spatial dimensions could also be
considered irrelevant if costs associated with accessing a behavior were equal,
e.g., if it was assumed to be equally easy for bumblebees to fly to alternative
flower ``species''~\cite{Baracchi2018} or for great tits to fly to different food
sources~\cite{Aplin2017}.  Each simulated environment provides a distinct challenge
to the agents, which will lead agents to differently evolve to become social
learners or not depending on which of those two strategies optimizes net payoffs.

\emph{Agents} here are autonomous, intelligent, simulated problem solvers. Each
agent evaluates its environment and chooses how to act within that environment based
on a set of empirically-motivated rules.  Our agents select which behavior to do
probabilistically, weighted by the softmax function applied to the mean payoff they
have observed for each behavior. The softmax function is a biologically plausible
generalization of behavior selection under uncertainty that enables agents to often
greedily exploit the most lucrative behavior (based on their observations) but also
to sometimes explore alternatives~\cite{Schulz2019,Collins2013,Daw2006,Yechiam2005}.
Such individual-level intelligence is a critical adaptation that has apparently
co-evolved with the capacity for social learning, and has a marked influence on the
cultural evolution of social learning as we will demonstrate in our Analysis.

In \emph{evolutionary} agent-based models like ours, agent life cycles are
programmed to cause agents to reproduce and die off, where reproducers pass on
heritable traits to their offspring. This sets up an inter-generational structure
that we also use as a scaffold for social learning: we assume social learning occurs
\emph{obliquely}\mt{is ``oblique learning'' understood to include parents by default
or do we need to say ``or vertically''?}, meaning learning happens when a new
generation of agents each select a teacher from the previous generation, possibly
including their parents.  Natural selection occurs in the model when agents with higher
payoffs preferentially reproduce. This is a model of cultural evolution, so
``reproduction'' means teaching the meta-cognitive strategy of whether one should learn
socially at all; ``death'', in this interpretation, means
loss of the potential to socially influence others. 
Evolutionary agent-based models are powerful, but their components need to be
carefully specified and justified in relation to empirical facts.

Evolution, then, acts as an optimization routine to identify which heritable trait
values are most rewarded under in different environmental uncertainty contexts,
defined by different uncertainty parameter settings
(Table~\ref{tab:uncertaintyParameters}).  The main outcome variable we analyze in
this model is therefore the value of agents' heritable trait, namely whether or not
an agent, $i$, learns socially, denoted $s_i$. $s_i$ is a binary variable: it is 1
if the agent is a social learner, and 0 otherwise. We will aggregate $s_i$ over all
agents and simulation trials to understand the affects of different uncertainty
parameter settings. We will support conclusions we make from observing patterns in
the outcomes with additional measures of the average time it takes for the
population to either all have $s_i = 1$ or all have $s_i = 0$, and the expected
payoffs in different contexts among populations that are initialized to all have
$s_i = 1$ or to all have $s_i = 0$. These outcome measures will be used to predict
the evolutionarily-favored strategy (to learn socially or not), the number of
generations it will take for a population to evolve the optimal strategy, and the
expected payoffs in populations performing the optimal strategy, in several specific
uncertainty contexts. To understand how different varieties of uncertainty interact
to affect the evolution of social learning, we systematically created an array of
simulated environments by systematically varying the uncertainty parameters.

% \begin{itemize}
%   \item
%     Softmax-guided behavior selection
%   \item
%     Oblique learning, not horizontal; no intragenerational change in optimal
%     behavior, only intergenerational
%   \item 
%     Payoff-biased teacher selection, not aggregation over everyone or large group,
%     as in conformist transmission
%   \item
%     Not spatial in any way. Agents need only to choose among potential behaviors
%     and the behavior is instantly performed, as a real human can when pointing
%     and clicking on a computer screen. This approximation could also hold when
%     all behavioral options require an identical amount of time/money/resources
%     to enable the agent to actually do the behavior (drive to work, walk to a
%     watering hole, fly to a flower to gather pollen, etc.). Since teacher and
%     behavior selection and reproduction are biased by relative payoffs there
%     is no effect on model dynamics if cost of access is the same for each behavior.
% \end{itemize}

\clearpage

\subsection{Model environment and uncertainty}

\begin{itemize}
  \item 
    A number of parameters specify the structure of the environment in which
    agents sovle problems and adapt by adjusting their social learning through
    evolution. \mt{Introduce all parameters; do not give values, but ref
    (Table~\ref{tab:uncertaintyParameters})}
  \item
    \mt{Bandits and payoffs in more detail}
\end{itemize}

\vspace{2em}
\begin{table}[h]
\caption{Uncertainty parameters.  Bold indicates default value tested} %\emph{default value tested}.}
    \label{tab:uncertaintyParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{4.0in}p{1.25in}} \toprule

        $B$       & Number of ``bandits'', i.e., behavior options
                  & 2, 4, 10 \\

        $L$    & Time steps per generation & 1, $B/2$, $B$, $2B$, $4B$ \\

        $N$    & Population size
                 & 50, 100, 200, \textbf{1000} \\

All agents perform individual, trial-and-error learning over their lifespan.
This learning is guided based on undirected, softmax search which drives
agents to explore all behavioral options in the selection set when 
the agent is more uncertain which is optimal, and to exploit the most profitable
behaviors when the agent is more certain which are most beneficial. Selection
for or against social learners occurs when agents are chosen to reproduce 
with replacement, with probability
of reproduction biased by net payoffs in each generation.

Several agent-level variables are required to operationalize individual and social
learning

\begin{table}[h]
  \caption{Agent-level variables. The first four ($\protect s_i$, $\protect
    \bar\pi_{ib}$, $\protect c_{ib}$,
  and $\protect \pi_i$) are dynamic with an implicit time dependence. The softmax
greediness $\protect \beta$ and number of prospective teachers for social learners,
$\protect N_T$, are constant throughout each simulation.}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{4.0in}p{1.25in}} \toprule

@@

        $\pi_i$ & Net payoffs accumulated by $i$ within generation & 0.0 \\

@@

        $N_T$    & Number of teachers to pool, from which best selected 
                 & 2, \textbf{5}, 10, 20  \\

@@

@@

\paragraph{Initialization}

@@

\subsubsection{Sensitivity analyses}

@@

By identifying four common uncertainty dimensions we developed a model that
predicts how different forms of uncertainty might interact to foster or suppress
social learning.  The model reproduced classic predictions that social learning can
yield higher payoffs than individual learning when uncertainty is sufficiently
limited.  By disambiguating and organizing various operationalizations of
uncertainty, we developed a more nuanced theoretical understanding of which forms
of uncertainty interact in which ways to make social learning more or less
beneficial compared to asocial learning. Taking the functional form of uncertainty
seriously is critical for understanding how humans adapt to an uncertain and
rapidly changing world, both to mitigate existential
threats~\cite{Moya2020,Jones2021}, and to capitalize on new opportunities such as
transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

Our model necessarily made some simplifying assumptions that may not always hold.
\mt{Review the ones mentioned in model section: non-spatial, no within-generation
environmental variability or horizontal learning}.
Furthermore, we only considered
success-biased learning, although conformity is theoretically important as
well~\cite{Muthukrishna2016a,Smaldino2018b}. One final implicit assumption
in our model was that the number of behaviors was constant, as was the relationship
between their payoffs---one was always optimal, and the rest were equally
suboptimal. However, this fails to account for evolutionary feedback which creates
new behavioral affordances as time progresses~\cite{Smaldino2012,Heras-Escribano2020}.
Our model and its software implementation were designed to be modular and 
therefore extensible.  Indeed, we plan to extend this model to further develop 
a more detailed theoretical understanding of the evolution of social learning, 
and we hope others do, too.

As a final note, we believe 
our modeling approach may have applications to developing social artificial
intelligence, sepcifically multi-agent reinforcement learning
systems~\cite{Sandholm1996,Ndousse2021,Gronauer2022,Jaques2019}.
It seems to be implicitly assumed by current multi-agent reinforcement learning
algorithms that social learning
would always be beneficial. However, we have shown in our work that sometimes 
it is more efficient for autonomous social agents to solve problems without
reference to social information across a variety of uncertainty conditions. 
Our model agents are essentially software
robots, so our results should predict the benefit to real-world social robots
to either learn socially or asocially if the assumptions we make here are 
reasonably met or implemented in the social robot design.
Our model dynamics could thus be used as an
evolutionary algorithm to update AI agents' social learning strategy without
explicit knowledge of, or need to approximate, the underlying uncertainty structure.
A designer would need only to program the robots to go through evolutionary
cycles of simulated ``reproduction'' and ``death''.
Our model could also predict the expected benefit of social learning, which may be
inherently costly in artificial multi-agent learning systems, and how long the
translated evolutionary algorithm may take to converge.

\bibliographystyle{apacite}
\bibliography{/Users/mt/workspace/Writing/library.bib}
% \bibliography{this.bib}
